### 2024-08-16 cognitive-bias.qmd
20240816 - performance heuristic; anecdotal fallacy

As described [above](#sec-cognitiveBiasOverview), heuristics are mental shortcuts that people use to handle the overwhelming amount of information to process.
Three important heuristics used in judgment and prediction in the face of uncertainty include [@Tversky1974]:

- [availability heuristic](#sec-heuristicsAvailability)
- [representativeness heuristic](#sec-heuristicsRepresentativeness)
- [anchoring and adjustment heuristic](#sec-heuristicsAnchoringAdjustment)

@@

@@

- [base rate fallacy](#sec-fallaciesBaseRate) (aka base rate neglect)
- [regression fallacy](#sec-fallaciesRegression)
- [hot hand fallacy](#sec-fallaciesHotHand)
- [sunk cost fallacy](#sec-fallaciesSunkCost)
- [gambler's fallacy](#sec-fallaciesGambler)
- [conditional probability fallacy](#sec-fallaciesConditionalProbability)
- [ecological fallacy](#sec-fallaciesEcological)

@@

@@

The ecological fallacy is the error of drawing inferences about an individual from group-level data.
For instance, a manager is interested in how age is associated with fantasy performance.
They examine the correlation between age and fantasy points and find that age is positively correlated with fantasy points among Quarterbacks.
The ecological fallacy would then take this finding to infer that an individual Quarterback is likely to increase in their performance with age.
We describe the ecological fallacy more in @sec-ecologicalFallacy.

>>>>

As described [above](#sec-cognitiveBiasOverview), heuristics are mental shortcuts that people use to handle the overwhelming amount of information to process.
Important heuristics used in judgment and prediction in the face of uncertainty include [@Tversky1974; @Critcher2014]:

- [availability heuristic](#sec-heuristicsAvailability)
- [representativeness heuristic](#sec-heuristicsRepresentativeness)
- [anchoring and adjustment heuristic](#sec-heuristicsAnchoringAdjustment)
- [performance heuristic](#sec-heuristicsPerformance)

@@

### Performance Heuristic {#sec-heuristicsPerformance}

The *performance heuristic* refers to the tendency for people to predict their improvement in a task based on their previous performance [@Critcher2014].
Although past performance tends to be correlated with future performance, people tend to show regression to the mean, as described in @sec-fallaciesRegression.
Thus, people who show strong performance in their first few games may in many cases show worse performance in subsequent games.
Moreover, initial performance does not necessarily indicate the extent of one's future *improvement*.
For instance, if a manager has a successful first season, the performance heuristic may lead them to overestimate the extent to which they will improve in their skill or performance in the future.

@@

- [base rate fallacy](#sec-fallaciesBaseRate) (aka base rate neglect)
- [regression fallacy](#sec-fallaciesRegression)
- [hot hand fallacy](#sec-fallaciesHotHand)
- [sunk cost fallacy](#sec-fallaciesSunkCost)
- [gambler's fallacy](#sec-fallaciesGambler)
- [anecdotal fallacy](#sec-fallaciesAnecdotal)
- [conditional probability fallacy](#sec-fallaciesConditionalProbability)
- [ecological fallacy](#sec-fallaciesEcological)

@@

### Anecdotal Fallacy {#sec-fallaciesAnecdotal}

The *anecdotal fallacy* is the error of drawing inferences based on personal experience or singular examples rather than based on stronger forms of evidence.
Although people tend to find stories and anecdotes compelling, they are among the weakest forms of evidence, as described in @#sec-internalExternalValidityTradeoffs.
Consider that a fantasy football commentator writes an article about how they won their league by drafting a Quarterback and Wide Receiver by drafting with their first two picks, respectively.
The anecdotal fallacy might lead a manager who reads the article to believe that they are more likely to win their league if they follow a similar strategy, despite more compelling evidence otherwise.

@@

The *ecological fallacy* is the error of drawing inferences about an individual from group-level data.
For instance, a manager is interested in how age is associated with fantasy performance.
They examine the correlation between age and fantasy points and find that age is positively correlated with fantasy points among Quarterbacks.
The ecological fallacy would then take this finding to infer that an individual Quarterback is likely to increase in their performance with age.
We describe the ecological fallacy more in @sec-ecologicalFallacy.
